---
title: "Getting started with modelblueprint"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with modelblueprint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
library(modelblueprint)
```

## What is a modelblueprint?

A `modelblueprint` is an S7 object that wraps a fitted model alongside its
training data, pipeline functions, and deployment metadata. The goal is a
single object that carries everything needed to reproduce predictions,
run diagnostics, and save/restore the model.

```{r create}
mb <- modelblueprint(
  model              = glm(vs ~ wt + hp, data = mtcars, family = binomial),
  train              = mtcars,
  y_name             = "vs",
  model_display_name = "logistic_vs",
  deploy_notes       = "Logistic regression baseline"
)

print(mb)
```

## Predicting

`predict()` applies the full pipeline — pre-processing, feature engineering,
model prediction, and post-processing — and always returns a numeric vector.

```{r predict}
preds <- predict(mb, mtcars)
head(preds)
```

## One-way analysis

`one_way()` shows the exposure-weighted mean of the target across bins of a
feature, giving a quick view of the marginal relationship. Pass `predictions = TRUE`
to overlay the model's in-sample predictions as a lift chart.

```{r one-way}
one_way(mb, var = "wt")
one_way(mb, var = "wt", predictions = TRUE)   # lift chart
```

## Partial dependence plot

`pdp()` reveals the marginal effect of a feature on model output, controlling
for all other features by averaging predictions across the training data.

```{r pdp}
pdp(mb, var = "wt", bins = 8L)
```

## Model diagnostics

```{r diagnostics}
# Gains chart with Gini coefficient
gain(mb)

# Predicted vs observed calibration
pred_vs_obs(mb)

# Grouped residuals with loess trend
residuals_grouped(mb, exposure_per_bin = 5)
```

## Feature engineering pipeline

Use `feat_eng_fun` to apply transformations before prediction — the same
function is applied to both training data and new data at prediction time.

```{r feat-eng}
mb_fe <- modelblueprint(
  model        = lm(mpg ~ wt + wt2, data = transform(mtcars, wt2 = wt^2)),
  feat_eng_fun = function(df) transform(df, wt2 = wt^2),
  train        = mtcars,
  y_name       = "mpg"
)

head(predict(mb_fe, mtcars))
```

## Saving and loading

`savemb()` serialises the complete blueprint — model, data, and pipeline
functions — to a `.tar.gz` archive. `loadmb()` restores it exactly.

```{r save-load}
savemb(mb, path = tempdir(), filename = "logistic_vs")
mb2 <- loadmb(file.path(tempdir(), "logistic_vs.tar.gz"))

# Predictions are identical
all.equal(predict(mb, mtcars), predict(mb2, mtcars))
```

## Example blueprints

The package ships with example constructors for common model types. Those
backed by base-R run without extra packages:

```{r examples-base}
mb_lm  <- mb_lm_regression()
mb_glm <- mb_glm_binomial()

one_way(mb_lm,  var = "wt", predictions = TRUE)
pdp(mb_lm,  var = "hp")
gain(mb_glm)
```

Constructors for tree-based and boosted models are also available but require
their respective packages to be installed:

```{r examples-optional, eval = requireNamespace("randomForest", quietly = TRUE) && requireNamespace("xgboost", quietly = TRUE)}
mb_rf  <- mb_rf_regression()      # requires randomForest
mb_xgb <- mb_xgb_classification() # requires xgboost

one_way(mb_rf,  var = "wt", predictions = TRUE)
pdp(mb_xgb, var = "hp")
```
