---
title: "Getting started with modelblueprint"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with modelblueprint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
library(modelblueprint)
```

## What is a modelblueprint?

A `modelblueprint` is an S7 object that wraps a fitted model alongside its
training data, pipeline functions, and deployment metadata. The goal is a
single object that carries everything needed to reproduce predictions,
run diagnostics, and save/restore the model.

The example below fits a Poisson frequency model on a small synthetic
car insurance dataset. `claim_freq` (claims per unit exposure) is the target;
`nclaims` is the raw count modelled with a log-exposure offset.

```{r create}
# Synthetic car insurance data ------------------------------------------------
set.seed(42L)
n   <- 500L
ins <- data.frame(
  driver_age  = sample(18L:75L, n, replace = TRUE),
  vehicle_age = sample(0L:15L,  n, replace = TRUE),
  gender      = sample(c("M", "F"), n, replace = TRUE),
  exposure    = round(runif(n, 0.1, 1.0), 2)
)
ins$nclaims    <- rpois(n, lambda = ins$exposure * exp(
  log(0.08) + ifelse(ins$driver_age < 25, 0.5, 0) + 0.01 * ins$vehicle_age
))
ins$claim_freq <- ins$nclaims / ins$exposure
train <- ins[1:400, ]
test  <- ins[401:500, ]

# Wrap the fitted model -------------------------------------------------------
mb <- modelblueprint(
  model = glm(
    nclaims ~ driver_age + vehicle_age + gender + offset(log(exposure)),
    data   = train,
    family = poisson
  ),
  post_process_fun   = function(preds, df_raw) preds / df_raw$exposure,
  train              = train,
  test               = test,
  y_name             = "claim_freq",
  expo_name          = "exposure",
  x_original_inputs  = c("driver_age", "vehicle_age", "gender"),
  model_display_name = "glm_poisson_freq",
  deploy_notes       = "GLM Poisson frequency: motor third-party liability"
)

print(mb)
```

## Predicting

`predict()` applies the full pipeline — pre-processing, feature engineering,
model prediction, and post-processing — and always returns a numeric vector.
The `post_process_fun` above converts predicted counts to predicted claim
frequency automatically.

```{r predict}
preds <- predict(mb, test)
head(preds)
```

## One-way analysis

`one_way()` shows the exposure-weighted mean of the target across bins of a
feature, giving a quick view of the marginal relationship. Pass
`predictions = TRUE` to overlay the model's in-sample predictions as a lift
chart.

```{r one-way}
one_way(mb, var = "driver_age", bins = 5)
one_way(mb, var = "driver_age", predictions = TRUE, bins = 5)   # lift chart
```

## Partial dependence plot

`pdp()` reveals the marginal effect of a feature on model output, controlling
for all other features by averaging predictions across the training data.

```{r pdp}
pdp(mb, var = "driver_age", bins = 10)
```

## Model diagnostics

```{r diagnostics}
# Each diagnostic defaults to one chart per available set (train/test/holdout);
# pass set = to pick one.

# Gains chart with Gini coefficient
gain(mb, set = "train")

# Predicted vs observed calibration
pred_vs_obs(mb, set = "train")

# Grouped residuals with loess trend
residuals_grouped(mb, set = "train", exposure_per_bin = 5)
```

Instead of calling each diagnostic individually, `model_validation()` runs
the full suite — validation, one-way, stability, PDP and SHAP plots for every
available set — and writes the results as structured HTML files:

```{r model-validation, eval = FALSE}
model_validation(mb, filepath = tempdir())
```

See `vignette("model-diagnostics")` for details on the individual charts and
the output layout.

## Feature engineering pipeline

Use `feat_eng_fun` to apply transformations before prediction — the same
function is applied to both training data and new data at prediction time.

```{r feat-eng}
mb_fe <- modelblueprint(
  model = glm(
    nclaims ~ driver_age + driver_age2 + vehicle_age + offset(log(exposure)),
    data   = transform(train, driver_age2 = driver_age^2),
    family = poisson
  ),
  feat_eng_fun       = function(df) transform(df, driver_age2 = driver_age^2),
  post_process_fun   = function(preds, df_raw) preds / df_raw$exposure,
  train              = train,
  y_name             = "claim_freq",
  expo_name          = "exposure"
)

head(predict(mb_fe, test))
```

## Saving and loading

`savemb()` serialises the complete blueprint — model, data, and pipeline
functions — to a `.tar.gz` archive. `loadmb()` restores it exactly.

```{r save-load}
savemb(mb, path = tempdir(), filename = "glm_poisson_freq")
mb2 <- loadmb(file.path(tempdir(), "glm_poisson_freq.tar.gz"))

# Predictions are identical
all.equal(predict(mb, test), predict(mb2, test))
```
