---
title: "Model diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
library(ModelBlueprint)
```

## Overview

ModelBlueprint provides four diagnostic charts for evaluating model
performance. Each works on a plain data frame or dispatches automatically
from a `ModelBlueprint` object.

| Function | Purpose |
|---|---|
| `gain()` | Rank-ordering ability — Gini coefficient |
| `pred_vs_obs()` | Calibration — are predicted rates accurate? |
| `residuals_grouped()` | Systematic bias across the prediction range |
| `sami()` | Where do two models disagree? |

---

## Gains chart (`gain`)

The gains chart measures how well a model **rank-orders** risk. Predictions
are sorted from highest to lowest, and the chart shows what proportion of
total observed losses is captured as we move through the portfolio.

A perfect model captures all losses in the top-risk segment. A random model
follows the diagonal. The **Gini coefficient** summarises rank-ordering ability
as a single number between -1 and 1 — higher is better.

```{r gain, eval = FALSE}
mb <- mb_glm_binomial()
mtcars$pred_model1 <- predict(mb, mtcars) + rnorm(nrow(mtcars))
mtcars$pred_model2 <- predict(mb, mtcars) + rnorm(nrow(mtcars))

# Gains chart — Gini shown in legend
gain(mb)

# Compare two models on the same chart
gain(
  mtcars,
  pred     = c("pred_model1", "pred_model2"),
  obs      = "vs"
)
```

**When to use it:** The gains chart is the primary performance metric in
insurance pricing. A Gini above 0.3 is generally considered useful for
pricing; above 0.5 is strong.

---

## Predicted vs observed (`pred_vs_obs`)

The calibration chart bins predictions and plots the average predicted rate
against the average observed rate per bin. A well-calibrated model has points
close to the diagonal — predicted and observed are equal on average.

```{r pred-vs-obs, eval = FALSE}
mb <- mb_glm_binomial()

# Calibration chart
pred_vs_obs(mb, ret = "data")

# More bins for finer resolution
pred_vs_obs(mb, bins = 20L)

# Equal-range bins (useful when predictions are skewed)
pred_vs_obs(mb, type_agg = "equal_range")
```

The chart has three elements:

- **Yellow bars** — exposure per bin (right axis)
- **Blue dots** — observed rate per bin
- **Dashed black line** — predicted rate per bin

When the dots track the dashed line closely, the model is well-calibrated.
Systematic divergence at the tails indicates the model underestimates or
overestimates extreme risks.

**When to use it:** Use this after fitting to check that predicted rates
match observed rates across the full prediction range. Also useful for
comparing a recalibrated vs uncalibrated model.

---

## Grouped residuals (`residuals_grouped`)

The residuals chart bins predictions by exposure and plots grouped residuals —
the difference between observed and predicted — with a loess trend line and
95% confidence interval.

A well-specified model has residuals scattered randomly around zero. A
systematic pattern indicates model misspecification.

```{r residuals, eval = FALSE}
mb <- mb_glm_binomial()

# Raw residuals (obs - pred)
residuals_grouped(mb)

# Pearson residuals — scaled by sqrt(pred), useful for count/rate models
residuals_grouped(mb, residual_type = "pearson")

# Control grouping granularity via exposure per bin
residuals_grouped(mb, exposure_per_bin = 500)
```

The `exposure_per_bin` argument controls how many data points go into each
group — smaller values give more groups and a noisier but more detailed view.

**When to use it:** Use this to diagnose where a model systematically
over- or under-predicts. An upward trend at the right tail means the model
underestimates high-risk predictions; a downward trend means it overestimates.

---

## SAMI double lift chart (`sami`)

The SAMI chart compares two competing models by binning the **ratio** of one
model's predictions to another. For each bin of the ratio, it shows the
observed mean alongside both model predictions.

Where the ratio is far from 1, the models disagree substantially. The
observed line reveals which model is closer to reality in that region.

```{r sami, eval = FALSE}
# Compare two blueprints
mb1 <- mb_lm_regression()
mb2 <- mb_glm_regression()

sami(list(mb1, mb2), bins = 10L)
```

On a plain data frame with predictions already attached:

```{r sami-df, eval = FALSE}
df        <- mtcars
df$pred1  <- predict(mb1, mtcars) + rnorm(nrow(mtcars))
df$pred2  <- predict(mb2, mtcars) + rnorm(nrow(mtcars))

sami(df, obs = "mpg", pred = c("pred1", "pred2"),
     exposure = "exposure", bins = 5, recalib = TRUE)
```

With `recalib = TRUE`, both predictions are scaled to match the observed
mean before computing ratios — this isolates shape differences from overall
level differences.

```{r sami-recalib, eval = FALSE}
sami(list(mb1, mb2), bins = 10L, recalib = TRUE)
```

**When to use it:** SAMI is most useful when deciding between two candidate
models. It shows not just which model performs better overall, but **where**
one model is better and by how much. The regions where the observed line
tracks one model more closely than the other drive the decision.

---

## Choosing the right diagnostic

```{r decision, echo = FALSE}
knitr::kable(data.frame(
  Question = c(
    "How well does the model rank-order risk?",
    "Are predicted rates accurate?",
    "Does the model have systematic bias?",
    "Which of two models is better, and where?"
  ),
  Diagnostic = c(
    "gain()",
    "pred_vs_obs()",
    "residuals_grouped()",
    "sami()"
  ),
  `Key metric` = c(
    "Gini coefficient",
    "Predicted vs observed alignment",
    "Residual pattern around zero",
    "Ratio of predictions vs observed"
  ),
  check.names = FALSE
))
```
