---
title: "Partial dependence plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Partial dependence plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
library(modelblueprint)
```

```{r mb, include = FALSE}
mb <- mb_glm_poisson_freq()
```

## What is a partial dependence plot?

A partial dependence plot (PDP) reveals the **marginal effect** of a single
feature on model output, averaging out the influence of all other features.
Unlike a one-way plot — which shows raw observed and predicted means per bin
— a PDP fixes a feature at a series of values and averages predictions across
the full training distribution.

This makes PDPs especially useful for:

- Understanding non-linear feature effects in black-box models
- Comparing the shape of a relationship across competing models
- Diagnosing interactions and unexpected behaviour

The chart has the same dual-axis layout as one-way plots:

- **Yellow bars (right axis)** — exposure per bin
- **Three lines (left axis)** — observed mean, in-sample predicted mean, and
  PDP (marginal effect)

## Basic usage

```{r basic}
# PDP for driver_age — everything pulled from blueprint slots
pdp(mb, var = "driver_age")
```

On a plain data frame:

```{r dataframe, eval = FALSE}
model <- glm(
  nclaims ~ driver_age + vehicle_age + offset(log(exposure)),
  data = mb@train, family = poisson
)
pdp(mb@train, var = "driver_age", obs = "claim_freq", model = model)
```

## Understanding the three lines

The PDP chart shows three lines per feature:

- **Observed** — exposure-weighted mean of the actual target per bin.
  This is the same as the one-way plot.
- **Predicted (in-sample)** — exposure-weighted mean of `predict(model, data)`
  per bin. Tracks the observed line when the model is well-calibrated.
- **PDP** — the true marginal effect. For each bin midpoint, the feature is
  fixed at that value across **all rows** of the training data and predictions
  are averaged. This isolates the feature's contribution from correlations
  with other features.

When the PDP line diverges from the predicted line, it signals that the
feature's in-sample predictions are driven partly by correlations rather
than the feature itself.

## Controlling bins and sample size

```{r bins}
# Coarser view
pdp(mb, var = "driver_age", bins = 8L)
```

PDP computation requires predicting across the full dataset for each bin.
For large datasets, `sample_size` caps the number of rows used:

```{r sample-size}
# Use at most 1,000 rows for PDP computation
pdp(mb, var = "vehicle_value", sample_size = 1000L)
```

The default is 10,000. For most models, 1,000 rows gives a stable PDP line
at a fraction of the computation cost.

## Binning strategy

```{r type-agg}
pdp(mb, var = "vehicle_age", type_agg = "equal_range")
```

## Choosing the dataset

```{r set}
pdp(mb, var = "driver_age", set = "train")
pdp(mb, var = "driver_age", set = "test")    # out-of-sample PDP
```

Computing the PDP on the test set is a useful check — if the PDP shape
changes substantially from train to test, the model may be overfitting to
correlations present only in training data.

## Returning the data

```{r ret-data}
d <- pdp(mb, var = "driver_age", ret = "data")
head(d)
```

Columns returned:

| Column | Description |
|---|---|
| `driver_age` | Bin label (interval string or level name) |
| `obs_mean` | Exposure-weighted observed mean |
| `pred_mean` | Exposure-weighted predicted mean |
| `pdp_mean` | Marginal PDP value |
| `exposure` | Total exposure in bin |
| `global_obs` | Dataset-wide observed mean |
| `global_pred` | Dataset-wide predicted mean |

## Compatible model types

`pdp()` works with any model that implements `predict()`:

```{r models, eval = FALSE}
# Poisson frequency GLM
pdp(mb_glm_poisson_freq(), var = "driver_age")

# Random forest
pdp(mb_rf_regression(), var = "wt")

# XGBoost
pdp(mb_xgb_regression(), var = "wt")

# H2O
pdp(mb_h2o_regression(), var = "wt")
```

For H2O models, `pdp()` connects to the running H2O cluster automatically
and handles the H2OFrame conversion internally.

## Interpreting PDPs

A flat PDP line means the feature has little marginal effect — the model
ignores it once other features are controlled for. A sloping line means the
feature drives predictions in that direction across the full data distribution,
not just where it correlates with other predictors.

For frequency models, PDP values are on the claim frequency scale. A PDP
that rises from 0.05 to 0.20 across the `driver_age` range means the model
assigns substantially higher expected claim frequency to younger drivers,
even after controlling for vehicle age, gender, and other features.
